multi-variate

### 2023-04-03 code/analysis_code/analysis_README.Rmd
Added LM for Nutrient Data

This folder contains code for analyzing cleaned data sets and aids in visualization of parameters of interest.\
(1) The dust subfolder contains `daily_dust.qmd`, which looks at dust concentration over the daily time series. It also contains `dust_and_copies.qmd` which combine Vibrio enumeration with dust data.\
(2) The env sub-folder contains `env_par.qmd` that examines temperature and salinity over the time series.\
(3) The qPCR sub-folder contains `daily_qmd` that has data on Vibrio enumeration. This is also seen in the `dust_and_copies.qmd` in the dust subfolder \
(4) The stats folder contains `dust_copies_stats.qmd` that runs models on variables of interest. \

>>>>

This folder contains code for analyzing cleaned data sets and aids in visualization of parameters of interest.\
(1) The dust subfolder contains `daily_dust.qmd`, which looks at dust concentration over the daily time series. It also contains `dust_and_copies.qmd` which combine Vibrio enumeration with dust data.\
(2) The env sub-folder contains `env_par.qmd` that examines temperature and salinity over the time series.\
(3) The qPCR sub-folder contains `daily_qmd` that has data on Vibrio enumeration. This is also seen in the `dust_and_copies.qmd` in the dust subfolder \
(4) The stats folder contains:
- `1_distributiom.qmd`: Checks distribution of variables of interest
- `2_cca.qmd`: Performs cross-correlation analysis to test for lags in time series (between dust input and Vibrio growth)
- `3_uni_variate.qmd` : runs univariate models on variable and predictors of interest.
- `4_multi_variate.qmd`: runs multivariate models on variable of interest (copies_mL)

### 2023-04-03 code/analysis_code/stats/2_cca.qmd
Added LM for Nutrient Data

```{r, include = FALSE}
dust0<- read_csv(here("data", "processed_data","dust_master.csv")) #Different Dust attributes
data0<- read_csv(here("data", "processed_data", "tx_master.csv")) #Master Data
```

@@

```{r}
dust<- dust0 %>%
  filter(!is.na(date),
         date %in% as.Date("2022-07-04"):as.Date("2022-07-19")) %>%
  mutate(t7sum = log10(t7sum)) #Log transform Dust conc (as it is not normal)

@@

```{r}
master0<- 
  dust %>% full_join(data) %>%
  filter(hr_cst %in% "01") #While this is technically not for hour 1, this gets ride of all duplicates for the sake of analysis.
master<- master0[c(1:10,15,33)]

@@

### Make Site DF

**Need to include 7/5 and 7/6 for previous dust data (will be used for lag analysis)**

@@

# Uni-variate Models

**From previous literature, we know that Vibrio have responded to dust 24-48 hours after input. This can make analysis tricky, as the dust concentration of a specific day does not necessarily correlate with the Vibrio copies of that same date. We can use cross-correlation analysis to assess a lag in a time series and see where this lag is most significant.**

\*\*Dust (t7sum) was log transformed

## Cross-correlation:

**For this, we cannot have NAs so we choose the strict daily time series**

\***Remember that dust is log transformed**

```{r}
#Blind Oso
bo_ccf<- bo %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
ccf(bo_ccf$t7sum, bo_ccf$copies_mL) 
title("Blind Oso DustxCopies CCA", line =1)

@@

: \*[**One important thing to note**]{.ul}: Our data set contains dust data for all of 2022. Our copies_mL data only contains values from 7/7/22 --> 7/19/22. Therefore, these CCA were made with only this time series. This limits our analysis, as we have dust data prior that can explain copies_mL on the earlier dates 7/7 or 7/8. But we cannot incorporate this into the CCA as we have NA for copies_mL for those early days. Below we will run linear models that DO contain these prior dust days. And since we have identified a lag, we can shift the dust data to line up with the copies_mL that it corresponds with directly - and we no longer have NAs. But this ultiamtely means that the lag identified above may not be the best fit for our model below, so some additional work is needed that tests other lags in the code. Using a combination of visual and CCA would be helpful.

>>>>

```{r, include = FALSE}
dust0<- read_csv(here("data", "processed_data","dust_master.csv")) #Different Dust permutations
data0<- read_csv(here("data", "processed_data", "tx_master.csv")) #Master Data containing copies_mL
```

@@

```{r}
dust<- dust0 %>%
  filter(!is.na(date),
         date %in% as.Date("2022-07-04"):as.Date("2022-07-19")) %>%
  mutate(t7sum = log10(t7sum)) #Log transform Dust conc (as it is not normally distributed)

@@

```{r}
master0<- 
  dust %>% full_join(data) %>%
  filter(hr_cst %in% "01") #While this is technically not for hour 1, this gets ride of all duplicates for the sake of analysis.
master<- master0[c(1:9,14,32)]

@@

### Make data frames for the three sites

Because we are working with lags. We need to include dust data from a few days [prior]{.ul} to sampling, as the dust concentrations will have an impact on Vibrio growth 1-2 days later.

@@

# Cross-correlation Analysis

From previous literature, we know that Vibrio have responded to dust 24-48 hours after input. This can make analysis tricky, as the dust concentration of a specific day does not necessarily correlate with the Vibrio copies of that same date. We can use cross-correlation analysis to assess a lag in a time series and see where this lag is most significant.

-   For this, we cannot have NAs so we choose the strict daily time series (7/7 - 7/19). We will use the data frames made above for creating the appropriate lag later.

-   Remember that dust is log transformed

```{r}
#Blind Oso
bo_ccf<- bo %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19'))) 
ccf(bo_ccf$t7sum, bo_ccf$copies_mL) #CCA between dust and copies_mL
title("Blind Oso DustxCopies CCA", line =1)

@@

: \*[**One important thing to note**]{.ul}: Our data set contains dust data for all of 2022. Our copies_mL data only contains values from 7/7/22 --> 7/19/22. We cannot have NAs when running CCA, therefore, these CCA were made with only this time series. This limits our analysis, as we have dust data prior that can explain copies_mL on the earlier dates 7/7 or 7/8. But we cannot incorporate this into the CCA as we have NA for copies_mL for those early days. Below we will run linear models that DO contain these prior dust days. And since we have identified a lag, we can shift the dust data to line up with the copies_mL that it corresponds with directly - and we no longer have NAs. But this ultiamtely means that the lag identified above may not be the best fit for our model below, so some additional work is needed that tests other lags in the code. Using a combination of visualization and CCA would be helpful. For the guld site, although CCA identified a lag of 1 day being significant, we can see in our exploratory analysis, that there is a \~2 day lag.

<!--# May need to update this last statement after meeting with Erin. 1 day lag may be the best. -->

<!--# May need to include nutrient data for cca -->

### 2023-04-03 code/analysis_code/stats/4_multi_variate.qmd
Added LM for Nutrient Data

---
title: "env_and_copies_stats"
format:
  html:
    theme: default
---

@@

```{r}
library(tidyverse)
library(here)
library(Hmisc) #For lag function
library(tidymodels) #For modeling
library(ggpmisc) #For R2 equation on graph
library(dotwhisker) #Fpr visualization
```

@@

```{r}
dust0<- read_csv(here("data", "processed_data","dust_master.csv")) #Different Dust attributes
data0<- read_csv(here("data", "processed_data", "tx_master.csv")) #Master Data
```

@@

```{r}
dust<- dust0 %>%
  filter(!is.na(date),
         date %in% as.Date("2022-07-04"):as.Date("2022-07-19"))

data<- data0 %>%
  filter(date %in% as.Date("2022-07-04"):as.Date("2022-07-19"))
```

@@

master<- master %>%
  select(copies_mL, t7sum, site, date) 
```

# Check Distributions

### Copies per mL

```{r}

#Normal
master %>% 
  ggplot() + geom_histogram(
    aes(
      x = copies_mL)) +
  labs(title = "Non-Log")+ 
  facet_wrap(~site)

#Log-Transformed
master %>% 
  ggplot() + geom_histogram(
    aes(
      x = log(copies_mL))) +
  labs(title = "LOG-TRANS")+ 
  facet_wrap(~site)
```

**We will use the log-transformed data as it has a \~ normal distribution.**

## Log-transform Copies

```{r}
#master <- 
 # master %>%
 # mutate(copies_mL = log(master$copies_mL))
```

### Make Site DF

```{r}
bo<- master %>%
  filter(site %in% "Blind Oso"| date %in% c( as.Date("2022-07-05"), as.Date("2022-07-06")))

c2<- master %>%
  filter(site %in% "Canals"| date %in% c( as.Date("2022-07-05"), as.Date("2022-07-06")))

rd<- master %>%
  filter(site %in% "Gulf"| date %in% c(as.Date("2022-07-05"), as.Date("2022-07-06")))
```

# Uni-variate Models

**From previous literature, we know that Vibrio have responded to dust 24-48 hours after input. This can make analysis tricky, as the dust concentration of a specific day does not necessarily correlate with the Vibrio copies of that same date. We can use cross-correlation analysis to assess a lag in a time series and see where this lag is most significant.**

## Cross-correlation:

```{r}
#Blind Oso
bo_ccf<- bo %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
ccf(bo_ccf$t7sum, bo_ccf$copies_mL) 
title("Blind Oso DustxCopies CCA", line =1)

#Canals
c2_ccf<- c2 %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
ccf(c2_ccf$t7sum, c2_ccf$copies_mL)
title("Canals DustxCopies CCA", line=1)

#Gulf
rd_ccf<- rd %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
ccf(rd_ccf$t7sum, rd_ccf$copies_mL) 
title("Gulf DustxCopies CCA", line =1)
```

**A negative Lag value indicates that x is a predictor of y, or in this case, dust is a predictor of copies_mL- so we will be looking at these negative values. A positive ACF indicates a positive relationship between the two variables, where a negative ACF indicates a negative relationship. According to the ccf plots, BO has a significant lag relationship a value of -1. In other words, dust occurs, and one day later, we see a response in copies_mL. C2 has a lag at -2, and RD at -1**

## Blind Oso Site:

#### Add one day lag

```{r}
bo2<- bo%>% mutate(
  lag =Lag(bo$t7sum, shift  = 1)) %>% #Make lag from ccf
filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
```

#### Create a linear model

```{r}
#Create Recipe for Growth 
growth_recipe_bo<- recipe(copies_mL ~ lag, data = bo2)

#Set up linear model
lm_mod<- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

#Workflow that adds recipe to model
Growth_wflow<- 
  workflow() %>%
  add_model(lm_mod) %>%
  add_recipe(growth_recipe_bo)

#Use workflow to fit model to  data set
growth_fit_bo<- Growth_wflow %>%
  fit(data = bo2)

#View as Tibble 
growth_fit_bo %>%
  extract_fit_parsnip() %>%
  tidy()
```

```{r}
aug_test_bo <- augment(growth_fit_bo, bo2)
```

```{r}
rmse <- aug_test_bo %>% rmse(truth = copies_mL, .pred)
rsq <- aug_test_bo %>% rsq(truth = copies_mL, .pred)
rd_metrics<- full_join(rmse, rsq)
```

Approaches based on such selection criteria essentially try to guess how the model would perform if it were to be fit to new data, without actually trying to do it (in contrast to CV).

```{r}
bo_mod<- lm(copies_mL ~ lag, data = bo2)
AIC(bo_mod)
```

```{r}
res_bo<- resid(bo_mod)
plot(fitted(bo_mod), res_bo)
abline(0,0)
```

### Plot

```{r}
aug_test_bo %>% ggplot(aes(copies_mL, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (BO+1)")
```

## Canals:

#### Add two day lag

```{r}
c22<- c2%>% mutate(
  lag =Lag(c2$t7sum, shift  = 2)) %>% #Make lag from ccf
filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
```

#### Create a linear model

```{r}
#Create Recipe for Growth 
growth_recipe_c2<- recipe(copies_mL ~ lag, data = c22)

#Set up linear model
lm_mod<- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

#Workflow that adds recipe to model
Growth_wflow<- 
  workflow() %>%
  add_model(lm_mod) %>%
  add_recipe(growth_recipe_c2)

#Use workflow to fit model to  data set
growth_fit_c2<- Growth_wflow %>%
  fit(data = c22)

```{r}
aug_test_c2 <- augment(growth_fit_c2, c22)
```

```{r}
rmse <- aug_test_c2 %>% rmse(truth = copies_mL, .pred)
rsq <- aug_test_c2 %>% rsq(truth = copies_mL, .pred)
rd_metrics<- full_join(rmse, rsq)
```

```{r}
c2_mod<- lm(copies_mL ~ lag, data = c22)
AIC(c2_mod)
```

```{r}
res_c2<- resid(c2_mod)
plot(fitted(c2_mod), res_c2)
abline(0,0)
```

### Plot

```{r}
aug_test_c2 %>% ggplot(aes(copies_mL, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (C2+2)")
```

## Gulf Site:

#### Add two day lag

```{r}
rd2<- rd%>% mutate(
  lag =Lag(rd$t7sum, shift  = 2)) %>% #Make lag from ccf
filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
```

#### Create a linear model

```{r}
#Create Recipe for Growth 
growth_recipe_rd<- recipe(copies_mL ~ lag, data = rd2)

#Set up linear model
lm_mod<- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

#Workflow that adds recipe to model
Growth_wflow<- 
  workflow() %>%
  add_model(lm_mod) %>%
  add_recipe(growth_recipe_rd)

#Use workflow to fit model to  data set
growth_fit_rd<- Growth_wflow %>%
  fit(data = rd2)

```{r}
aug_test_rd <- augment(growth_fit_rd, rd2)
```

```{r}
rmse <- aug_test_rd %>% rmse(truth = copies_mL, .pred)
rsq <- aug_test_rd %>% rsq(truth = copies_mL, .pred)
rd_metrics<- full_join(rmse, rsq)
```

```{r}
rd_mod<- lm(copies_mL ~ lag, data = rd2)
AIC(rd_mod)
```

```{r}
res_rd<- resid(rd_mod)
plot(fitted(rd_mod), res_rd)
abline(0,0)
```

### Plot

```{r}
aug_test_rd %>% ggplot(aes(lag, copies_mL)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (RD+2)")
```

@@

#### Make New Dataset

```{r}
master2<- dust0 %>% full_join(data) %>%
  filter(hr_cst %in% "01") %>%
  select(site, sal, temp, amm, nn,orthop, copies_mL, t7sum) %>%
  na.omit()
```

#### Site-specific Data sets

```{r}
bo_a<- master2 %>%
  filter(site %in% "Blind Oso") %>%
  select(!site)

c2_a<- master2 %>%
  filter(site %in% "Canals")%>%
  select(!site)

rd_a<- master2 %>%
  filter(site %in% "Gulf")%>%
  select(!site)
```

## BO: Using all Variables vs. Growth

```{r}
#Create Recipe for Growth and all other parameters
growth_recipe_bo<- recipe(copies_mL ~ ., data = bo_a)

@@

```{r}
tidy(growth_fit_bo) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```
### Use model to test data

@@

```{r}
bo_a_mod<- lm(copies_mL ~ ., data = bo_a)
AIC(bo_a_mod)
```

@@

```{r}
res_bo_a<- resid(bo_a_mod)
plot(fitted(bo_a_mod), res_bo_a)
abline(0,0)
```
## C2: Using all Variables vs. Growth

@@

```{r}
tidy(growth_fit_c2) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```
### Use model to test data

@@

```{r}
res_c2_a<- resid(c2_a_mod)
plot(fitted(c2_a_mod), res_c2_a)
abline(0,0)
```
## RD: Using all Variables vs. Growth

@@

```{r}
tidy(growth_fit_rd) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```
### Use model to test data

>>>>

---
title: "multi-variate"
format:
  html:
    theme: default
---

@@

```{r, include = FALSE}
library(tidyverse)
library(here)
library(Hmisc) #For lag function
library(tidymodels) #For modeling
library(ggpmisc) #For R2 equation on graph
library(dotwhisker) #Fpr visualization
library(lubridate)#Dates
```

@@

```{r}
dust<- read_csv(here("data", "processed_data","dust_master.csv")) #Different Dust attributes
data<- read_csv(here("data", "processed_data", "tx_master.csv")) #Master Data
```

@@

```{r}
dust<- dust %>%
  filter(!is.na(date),
         date %in% as.Date("2022-07-04"):as.Date("2022-07-19"))

data<- data %>%
  filter(date %in% as.Date("2022-07-04"):as.Date("2022-07-19"))
```

@@

master<- master %>%
  select(site, sal, temp, copies_mL,t7sum, date) #%>%
 # na.omit()
```

@@

### Site-specific Data sets and Clean df

```{r}
bo_a<- master %>%
  filter(site %in% "Blind Oso" | date %in% as.Date('2022-07-05') : as.Date('2022-07-06'))
bo_a<- bo_a %>%
mutate(
  lag =Lag(bo_a$t7sum, shift  = 1)) %>% #Create 1 day lag
filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19'))) %>%
select(!c(date,t7sum,site))

c2_a<- master %>%
  filter(site %in% "Canals" | date %in% as.Date('2022-07-05') : as.Date('2022-07-06'))
c2_a<- c2_a %>%
mutate(
  lag =Lag(c2_a$t7sum, shift  = 2)) %>% #Make 2 day lag
filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))%>%
select(!c(date,t7sum,site))

rd_a<- master %>%
  filter(site %in% "Gulf" | date %in% as.Date('2022-07-05') : as.Date('2022-07-06')) 
rd_a<- rd_a %>%
mutate(
  lag =Lag(rd_a$t7sum, shift  = 2)) %>% #Make 2 day lag
filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))%>%
select(!c(date,t7sum,site))
```

## BO: Growth\~ Sal+Dust

```{r}
#Create Recipe for Growth and all other parameters
growth_recipe_bo<- recipe(copies_mL ~ sal+lag, data = bo_a)

@@

```{r}
tidy(growth_fit_bo) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```

@@

```{r}
bo_a_mod<- lm(copies_mL ~ temp+sal+lag, data = bo_a)
AIC(bo_a_mod)
```

@@

```{r}
res_bo_a<- resid(bo_a_mod)
plot(fitted(bo_a_mod), res_bo_a)
abline(0,0)
```

## C2: Using all Variables vs. Growth

@@

```{r}
tidy(growth_fit_c2) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```

@@

```{r}
res_c2_a<- resid(c2_a_mod)
plot(fitted(c2_a_mod), res_c2_a)
abline(0,0)
```

## RD: Using all Variables vs. Growth

@@

```{r}
tidy(growth_fit_rd) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```
